Solution-aware vs global ReLU selection: \\ partial MILP strikes back for DNN verification

### 2025-01-26 CAV 2025 TODO/Paper_noProof.tex
paper
fix
valabs

%\author{Antiquus S.~Hippocampus, Natalia Cerebro \& Amelie P. Amygdale \thanks{ Use footnote for providing further information
%		about author (webpage, alternative address)---\emph{not} for acknowledging
%		funding agencies.  Funding acknowledgements go at the end of the paper.} \\
%	Department of Computer Science\\
%	Cranberry-Lemon University\\
%	Pittsburgh, PA 15213, USA \\
%	\texttt{\{hippo,brain,jen\}@cs.cranberry-lemon.edu} \\
%	\And
%	Ji Q. Ren \& Yevgeny LeNet \\
%	Department of Computational Neuroscience \\
%	University of the Witwatersrand \\
%	Joburg, South Africa \\
%	\texttt{\{robot,net\}@wits.ac.za} \\
%	\AND
%	Coauthor \\
%	Affiliation \\
%	Address \\
%	\texttt{email}
%}

@@

%\input{valabstraction}

@@

{\bf Reproducibility Statement:} 
We tested twice outlier results to confirm them, making sure of reproducibility on the given hardware. Precise details on the settings used are provided in the appendix.
Additional results {\color{blue}(e.g. ablation studies)} are also provided in the appendix.
Tested DNNs as well as MNIST and CIFAR10 DataSet are freely available.
The source code of Hybrid MILP will be provided on GitHub after acceptance (needing Gurobi as well as $\alpha,\beta$-Crown).

>>>>

@@

\input{valabstraction}

@@

%{\bf Reproducibility Statement:} We tested twice outlier results to confirm them, making sure of reproducibility on the given hardware. Precise details on the settings used are provided in the appendix. Additional results {\color{blue}(e.g. ablation studies)} are also provided in the appendix. Tested DNNs as well as MNIST and CIFAR10 DataSet are freely available. The source code of Hybrid MILP will be provided on GitHub after acceptance (needing Gurobi as well as $\alpha,\beta$-Crown).

### 2025-01-26 CAV 2025 TODO/valabstraction.tex
valabs

\subsection{MILP, LP and partial MILP encodings for DNNs}

@@

MILP instances can be linearly relaxed into LP over-abstraction, where variables originally restricted to integers in $\{0,1\}$ (binary) are relaxed to real numbers in the interval $[0,1]$, while maintaining the same encoding. As solving LP instances is polynomial time, this optimization is significantly more efficient. However, this efficiency comes at the cost of precision, often resulting in less stringent bounds. This approach is termed the {\em LP abstraction}.

In this paper, we propose to use {\em partial MILP}, to get interesting trade-offs between accuracy and runtime: 
for a set of unstable neurons $X$, we denote by MILP$_X$ the MILP encoding where variables encoding $X$ are binary, and other variables are linear variables using the LP relaxation. We say that nodes in $X$ are {\em opened}. 
To further limit the number of binary variables needed for a given accuracy, we devise the same iterative approach as the box abstraction or DeepPoly \cite{deeppoly}, computing lower 
and upper bounds $\LB(n),\UB(n)$ for neurons $n$ of a layer, that are used when computing values of the next layer, thus necessitating less variables from previous layers.

The crucial factor in such an approach is to {\em select} few opened ReLU nodes in $X$ which are the most important for the accuracy. An extreme strategy was adopted in \cite{DivideAndSlide}, where only ReLU nodes of the immediate previous layer can be opened, and the measure to choose ReLU $a$ when computing the bounds for neuron $b$ was to consider $|W_{ab}| (\UB(a)-\LB(a))$. 
To obtain a more accurate measure, that 
%can also scale to deeper layers and 
is not limited to open nodes from the immediate previous layer, 
{\color{blue} we invoke a folklore result \cite{alessandro} on the LP relaxation of (\ref{eq11}), for which we provide a direct and explicit proof:}

@@

\begin{proposition}
	\label{LP}
	The LP relaxation of (\ref{eq11}) is equivalent with the triangular abstraction 	(\ref{eq:deeppoly}).
\end{proposition}

@@

>>>>

\subsection{MILP, and LP encodings for DNNs}

@@

MILP instances can be linearly relaxed into LP over-abstraction, where variables originally restricted to integers in $\{0,1\}$ (binary) are relaxed to real numbers in the interval $[0,1]$, while maintaining the same encoding. As solving LP instances is polynomial time, this optimization is significantly more efficient. However, this efficiency comes at the cost of precision, often resulting in less stringent bounds. This approach is termed the {\em LP abstraction}. We invoke a folklore result on the LP relaxation of (\ref{eq11}), for which we provide a direct and explicit proof:

@@

\begin{proposition}
	\cite{alessandro}
	\label{LP}
	The LP relaxation of (\ref{eq11}) is equivalent with the triangular abstraction 	(\ref{eq:deeppoly}).
\end{proposition}

@@

\section{Partial MILP}

In this paper, we revisit the use of {\em partial MILP} (see \cite{DivideAndSlide}), to get interesting trade-offs between accuracy and runtime: 
for a set of unstable neurons $X$, we denote by MILP$_X$ the MILP encoding where variables encoding $X$ are binary, and other variables are linear variables using the LP relaxation. We say that nodes in $X$ are {\em opened}. To further limit the number of binary variables needed for a given accuracy, we devise the same iterative approach as the box abstraction or DeepPoly \cite{deeppoly}, computing lower and upper bounds $\LB(n),\UB(n)$ for neurons $n$ of a layer, that are used when computing values of the next layer, thus necessitating less variables from previous layers.

%To obtain a more accurate measure, that 
%can also scale to deeper layers and 
%is not limited to open nodes from the immediate previous layer,

\SetKwInput{KwInput}{Input}
\SetKwInput{KwOutput}{Output}

\begin{algorithm}[h!]
	\caption{pMILP($K$)}
	\label{algo1}
	\KwInput{Bounds $[\alpha_n,\beta_n]$ for input nodes $n$ at layer $0$ (input neighbourhood)}

	\KwOutput{Bounds $[\alpha_n,\beta_n]$ for every node $n$}

	\For{layer $k=1 \cdots \ell$}{
		\For{neuron $n$ in layer $k$}{

			Compute $X$ a set of $K$ nodes with the highest Utility for target neuron $n$.

			Run MILP$_X$ to obtain $[\alpha_n,\beta_n]$ from bounds of neurons in layers $< k$
		}
	}
\end{algorithm}

%We then analyse its complexity.
%First, notice that the complexity to rnu Utility is negligible wrt the Run MILP$_X$.

We provide the pseudo code for pMILP in Algorithm \ref{algo1}.
pMILP($K$) has a worst case complexity bounded by $O(N \cdot \MILP(N,K))$, 
where $N$ is the number of nodes of the DNN, and $\MILP(N,K)$ is the complexity of solving a MILP program with $K$ integer variables and $N$ linear variables.
We have $\MILP(N,K) \leq 2^K \LP(N)$ where $\LP(N)$ is the Polynomial time to solve a Linear Program with $N$ variables, $2^K$ being an upper bound. Solvers like Gurobi are quite adept and usually do not need to evaluate all $2^K$ ReLU configurations to deduce the bounds.
It is worth mentioning that the "for" loop iterating over neurons $n$ in layer $k$ (line 2) can be executed in parallel, because the computation only depends on bounds from preceding layers, not the current layer $k$. This worst-case complexity compares favorably when $K<<N$ with the worst-case complexity of BaB which would be $O(out \times 2^N)$, where $out$ is the number of output neurons. Of course, actual complexity is always much better than the worst case, thanks to different heuristics, such as pruning, which is what happens for easy instances, for which the linear cost $N$ in pMILP (which cannot be avoided) is actually the limiting factor. Only experimental results can tell whether the accuracy lost using reduced variables in pMILP can be a good trade off vs efficiency, both depending on the number $K$ chosen.
%This parallelization results in a time complexity of $O(\sum_{i=1}^{\ell} \MILP(N_{i-1},K))$, where $\ell$ denotes the total number of layers and $N_i$ the count of neurons in layers $0, ..., i$. 
%pMILP uses 20 threads.

If $K$ is sufficiently small, 
%Algorithm \ref{algo1} invokes many MILP (e.g., using Gurobi) with a manageable number of %integer variables. 
this approach is expected to be efficient. The crucial part is thus to find few neurons which are particularly important when computing neuron $n$. This is where our novel Utility function outperforms previous solutions, by using an call to an (LP) solver to obtain a solution, which is novel as far as we know. This solutions allows to optimizing the choice of open nodes for the particular input considers, and appears significantly better than previous attempts, explaining the efficiency of the method.

%For comparison, refined $\beta$-Crown and refined Prima used all the nodes up to a certain layer as binary variables, which is particularly inefficient, see Table \ref{table12} and Fig. \ref{fig4}. Hence, it can only be applied to small DNNs (it is implemented only up to MNNIST 8x200), while pMILP scales to at least 20.000 neurons (CNN-B-Adv).\cite{DivideAndSlide} is the closest to pMILP, also implementing a choice of nodes as binary variables. However, their choice is particularly inefficient, as revealed by Table \ref{tab:example1}, needing 4 times the number of open nodes for the same accuracy as our Utility function.

The crucial factor in such an approach is to {\em select} few opened ReLU nodes in $X$ which are the most important for the accuracy. An extreme strategy was adopted in \cite{DivideAndSlide}, where only ReLU nodes of the immediate previous layer can be opened, and the measure to choose ReLU $a$ when computing the bounds for neuron $b$ was to consider $|W_{ab}| (\UB(a)-\LB(a))$.
